Hydras \& Co.

### 2023-11-15 doc/chapter-primrec.tex
Corrections to chapter 11 (coq and doc)

\section{Introduction}
Primitive recursive functions are a small class of total functions, corresponding to the expressive power of a simple imperative programming language without \textbf{while} loops, in which every program execution terminates.

\emph{Primitive recursive functions} are total functions from $\mathbb{N}^n$ to
$\mathbb{N}$, for some $n\in\mathbb{N}$. Note that not all 
total $n$-ary recursive functions are primitive recursive
(see for instance Sect.~\vref{sect:ack-not-PR}).

@@

The traditional definition of the set of primitive recursive functions is structured as an inductive definition 
in five rules: three base cases, and two recursive construction rules.

@@

\subsection{A few (informal) examples}
Before playing with primitive recursive functions in \coq{}, let us get familiar with the definition above, with the help of a few simple examples.
\label{sect:primrec-examples1}

@@

Let $k$ be some natural number; the unary constant function which always returns $k$ is built through $k$ nested compositions of the 
successor function to the  unary constant function which returns $0$.

@@

\begin{align*}
  0 + x_1 &:= x_1 \\
  S x_0 + x_1 &:= S (x_0 + x_1)
\end{align*}

@@

\begin{remark}

The four examples above are defined as \coq terms in Section~\vref{sect:primrec-examples2}

\end{remark}
  We present in this chapter a formalization of  primitive recursive functions, taken from  Russel O'Connor's formalization in \coq{} of
G\"odel's incompleteness theorems~\cite{OConnor05}.

This chapter contains some comments on Russel's library, as well as a few extensions.
Contributions (under the form of comments, new examples or exercises) are welcome!.

@@

@@

\section{Basic definitions}
\index{maths}{Primitive recursive functions}

@@

The  \texttt{primRec} library allows us to consider primitive functions on \texttt{nat}, with any number of arguments, in 
curried form. This is made possible in 
\href{../theories/html/hydras.Ackermann.extEqualNat.html}{Ackermann.extEqualNat} by the following definition:

@@

Likewise, arbitrary boolean predicates may have an arbitrary number of arguments. The dependent type
(\texttt{naryRel $n$}), defined in the same way as \texttt{naryFunc}, is the type of $n$-ary functions from
\texttt{nat} into \texttt{bool}.

\input{movies/snippets/PrimRecExamples/naryRel2}

The magic of dependent types makes it possible to define recursively extensional equality between functions of the same arity.

@@

\index{ackermann}{Predicates!extEqual}

\input{movies/snippets/extEqualNat/extEqualDef}

@@

@@

\index{coq}{Dependent types}
\index{coq}{Mutually inductive types}

@@

\subsection{A few arithmetic functions}
\label{sect:primrec-examples2}
Let us show how the few functions presented in Section~\vref{sect:primrec-examples2} can be defined in \coq.

@@

\subsection{A little bit of semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$.
This function is defined by mutual recursion,  together with the  function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

@@

\index{coq}{Dependent pattern matching}
Both functions are mutually defined through dependent pattern matching. We advise the readers who 
would feel uneasy with dependent types to consult Adam Chlipala's \emph{cpdt}  book~\cite{chlipalacpdt2011}. We leave it to the reader  to look also at the helper functions in
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.

@@

Looks complicated? The following examples show that, when
the arity is fixed, these definitions behave well w.r.t. 
\coq's reduction rules. Moreover, they make the interpretation functions more ``concrete''.

\subsubsection{A few tests}
Let us consider again the functions defined in 
previous sections.

@@

\subsubsection{A few proofs of correctness}

It is tempting to \emph{prove} that our functions \texttt{cst0},
\texttt{plus}, \texttt{mult} and \texttt{fact} are correct implementations in \texttt{PrimRec} of the mathematical functions we
consider.

@@

\section{Proving that a given arithmetic function is primitive recursive}
\label{sect:proofs-of-isPR}

The example in the preceding section clearly shows that, in order to prove that a given arithmetic function
(defined in \gallina{} as usual) is primitive recursive, trying to give  by hand a term  of type (\texttt{PrimRec $n$}) is not a good method, since such terms may be huge and complex, even for simple arithmetic functions. The method proposed in Library \texttt{primRec} is the following one:

\begin{enumerate}
\item Define a type corresponding to the statement "the function \texttt{$f$:naryFunc $n$} is primitive recursive ''.
\item Prove handy lemmas which may help to prove that a given function is primitive recursive.
\end{enumerate}

Thus, the proof that a function, like \texttt{factorial}, is primitive recursive may be interactive, without having to type complex terms at any step of the development.

@@

The library \texttt{primRec} contains a large catalogue of lemmas allowing to prove statements 
of the form (\texttt{isPR $n$ $f$}). We won't list all these lemmas here, but give a few examples of
how they may be applied.

\begin{remark}
In the library \texttt{primRec}, all these lemmas are opaque (registered with \texttt{Qed}). Thus they do not allow the user to look at the witness of a proof of a \texttt{isPR} statement. Our example of page~\pageref{sect:bigfac} was built using a  copy of \texttt{primRec.v} where many \texttt{Qed}s have been replaced with
\texttt{Defined}s.

If it does not cause compatibility problems (with 
\href{https://github.com/coq-community/goedel}{goedel library} for instance), we plan to make all theses lemmas transparent.
\end{remark}

@@

The constructors \texttt{zeroFunc}, \texttt{succFunc},  and \texttt{projFunc} of type
\texttt{PrimRec} allows us to write trivial proofs of primitive recursivity. 
Some of   the following lemmas are already proven in 
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}. More examples are given in 
\href{../theories/html/hydras.MoreAck.PrimRecExamples.html}%
{Ackermann.MoreAck.PrimRecExamples.v}, in order to illustrate the main proof patterns.

@@

Likewise, we prove that the successor function on \texttt{nat} is primitive recursive too.

\input{movies/snippets/primRec/SuccIsPR}

@@

Please note that the projection $\pi_{1,1}$ is just the identity on \texttt{nat}, and is realized by 
(\texttt{projFunc 1 0}).

@@

@@

Let us look at the proof that any constant $n$ of type \texttt{nat} has type (\texttt{PR 0})
(lemma  \texttt{const1\_NIsPR} of \texttt{primRec}). We carry out a proof by induction on $n$, the base case of which is already proven.
Now, let us assume $n$ is \texttt{PR $n$}, with $x:\texttt{PrimRec}\,0$ as a ``realizer''.
Thus we would like to compose this constant function with the unary successor function.

@@

\vspace{4pt}
Thus, we get a quite simple proof of \texttt{const1\_NIsPR}.

@@

\subsubsection{Proving that \texttt{plus} is primitive recursive}

The lemma \texttt{plusIsPR} is already proven in \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}. We present in 
\href{../theories/html/hydras.MoreAck.PrimRecExamples.html}{MoreAck.PrimRecExamples}
a commented version of this proof,

First, we look for lemmas which may help to prove that a given function obtained with the recursor \texttt{nat\_rec} is primitive recursive.

\input{movies/snippets/PrimRecExamples/PrimRecExamplesSearch}

The following lemma shows it suffices to prove that
Standard library's function \texttt{plus} is extensionally equal to a function defined with
\texttt{nat\_rec}.

\input{movies/snippets/primRec/isPRExtEqual}

Thus, let us define an helper and prove its equivalence with \texttt{plus}.

\input{movies/snippets/PrimRecExamples/plusAlt}

@@

We are now able to complete the proof.

\input{movies/snippets/PrimRecExamples/plusIsPRa}

@@

\vspace{4pt}
\input{movies/snippets/PrimRecExamples/checkFilter0101IsPR}
\vspace{4pt}

Thus, our first subgoal is solved easily. The rest of the proof 
is just an application of already proven lemmas.

@@

\begin{todo}
Comment more examples from   \href{../theories/html/hydras.MoreAck.PrimRecExamples.html}{MoreAck.PrimRecExamples}.
\end{todo}

\index{ackermann}{Exercises}
\begin{exercise}
There is a lot of lemmas similar to \texttt{filter010IsPR} in the \texttt{primRec} library, useful to control the arity of functions.
Thus, the reader may look at them, and invent simple examples of application for each lemma.
\end{exercise}

\index{ackermann}{Exercises}
\begin{exercise}
Multiplication of natural number is already proven in the \texttt{primRec} library. Write a proof of your own, then compare to the library's version.
\end{exercise}

%\input{movies/snippets/PrimRecExamples/justATest}

@@

\input{movies/snippets/MorePRExamples/factDef}

@@

\index{ackermann}{Exercises}
\begin{exercise}
Show that the function \texttt{min: naryFunc\,2} is primitive
recursive.

\emph{You may start this exercise with
    \href{https://github.com/coq-community/hydra-battles/blob/master/exercises/primrec/MinPR.v}{exercises/primrec/MinPR.v}.}

>>>>

\section{Introduction}
\emph{Primitive recursive functions} are a small class of total arithmetic functions  from $\mathbb{N}^n$ to
$\mathbb{N}$, for some $n\in\mathbb{N}$, corresponding to the expressive power of a simple imperative programming language without \textbf{while} loops, in which every program execution terminates.
Please  note that not all 
total $n$-ary recursive functions are primitive recursive
(see for instance Sect.~\vref{sect:ack-not-PR}).

@@

The traditional definition of primitive recursive functions is structured as an inductive definition 
in five rules: three base cases, and two recursive construction rules.

@@

\subsection{A few (informal) examples}
\label{sect:PR-informal}
\label{sect:primrec-examples1}
Before playing with primitive recursive functions in \coq{}, let us get familiar with their mathematical definition, with the help of a few simple examples, which will be considered again
as \coq terms in Section~\vref{sect:primrec-examples2}.

@@

Let $k$ be some natural number; the unary constant function which always returns $k$ is built through $k$ nested compositions of the 
successor function with the  unary constant function which returns $0$.

@@

\begin{align*}
  0 + x_1 &:= x_1 \\
  S\; x_0 + x_1 &:= S (x_0 + x_1)
\end{align*}

@@

@@

  We present here a formalization of  primitive recursive functions, taken from  Russel O'Connor's formalization in \coq{} of
G\"odel's incompleteness theorems~\cite{OConnor05}.

A few additions and/or small changes (mainly notations, and adaptation to the continuously evolving practices of \coq development) have been made to
O'Connor's original library. 
Contributions (under the form of comments, new examples or exercises) are welcome!

@@

\section{Abstract syntax for primitive recursive functions}
\index{maths}{Primitive recursive functions}

@@

The  \href{../theories/html/hydras.Ackermann.extEqualNat.html}{Ackermann.extEqualNat}
 library allows us to consider primitive functions on \texttt{nat}, with any number of arguments, in curried form. This is made possible in 
by the following definition:

@@

\subsection{Extensional equality}

Dependent types make it possible to define recursively extensional equality between functions of the same arity.

@@

\index{ackermann}{Predicates!extEqual}
\inputsnippets{extEqualNat/extEqualDef}

Module \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec} defines and export  the notation ``\texttt{$f$ =x= $g$}'' for ``\texttt{extEqual $n$ $f$ $g$}'' \footnote{in parsing mode, provided $f$ has explicitely the type (\texttt{naryFunc $n$}).}

@@

\subsection{Boolean predicates}

Like arithmetic functions,  arbitrary boolean predicates may have an arbitrary number of arguments. The dependent type
(\texttt{naryRel $n$}), defined in the same way as \texttt{naryFunc}, is the type of $n$-ary functions from
\texttt{nat} into \texttt{bool}.

\noindent
\emph{From \href{../theories/html/hydras.Ackermann.extEqualNat.html}{Ackermann.extEqualNat}}
\inputsnippets{extEqualNat/naryRelDef}

\inputsnippets{primRec/ltBoolDef, primRec/leBoolDef}

\noindent
\emph{From \href{../theories/html/hydras.Ackermann.extEqualNat.html}{Ackermann.extEqualNat}}
\input{movies/snippets/PrimRecExamples/naryRel2}

@@

\begin{remark}
The \texttt{PrimRec}  type family is indeed a kind of 
programming language for writing primitive recursive functions.
The link to the mathematical notion of such functions will be
established in Section~\ref{primrec-semantics} when we give a semantics which maps any term of type (\texttt{PrimRec $n$}) 
to a function of type (\texttt{naryFunc $n$}).
\end{remark}
\index{coq}{Dependent types}
\index{coq}{Mutually inductive types}

@@

\subsubsection{Examples}
\label{sect:primrec-examples2}

Let us show how the functions described in \ref{sect:PR-informal} can be described by terms of type ``\texttt{PrimRec \_}''.

@@

\subsection{A little bit of semantics} 
\label{primrec-semantics}

Inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or \texttt{Arith.Factorial.fact}, etc. but terms of an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$.
This function is defined by mutual recursion,  together with the  function 
(\texttt{evalprimRecS $n$ $m$}) of type 
($\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$).

@@

\index{coq}{Dependent pattern matching}
Both functions are mutually defined through dependent pattern matching. We advise the readers who 
would feel uneasy with dependent types to consult Adam Chlipala's \emph{cpdt}  book~\cite{chlipalacpdt2011}. We invite the 
 reader  to look also at the helper functions in
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}, namely \texttt{evalConstFunc}, \texttt{evalProjFunc}, \texttt{evalComposeFunc}, and \texttt{evalPrimRecFunc}, etc.

@@

\subsubsection{A few tests}

The following examples show that the functions 
\texttt{evalPrimRec} and \texttt{evalPrimRecs}
 behave well w.r.t. 
\coq's reduction rules. They can also be considered as 
elementary tests of our definitions of \texttt{cst0}, \texttt{cst},
\texttt{plus}, \texttt{mult} and \texttt{fact}.

@@

\subsubsection{Correctness proofs}

It is now time to \emph{prove} that our functions \texttt{cst0},
\texttt{cst}, \texttt{plus}, \texttt{mult} and \texttt{fact} are correct implementations in \texttt{PrimRec} of the mathematical functions we
consider.

@@

\section{Proving that a given \coq arithmetic function is primitive recursive}
\label{sect:proofs-of-isPR}

The example in the preceding section clearly shows that, in order to prove that a given arithmetic function
(defined in \gallina{} as usual) is primitive recursive, trying to \emph{type} a term  of type (\texttt{PrimRec $n$}) is not a good method, since such terms may be too large, even for simple arithmetic functions. The method proposed in the library \texttt{Ackermann.primRec} is the following one:

\begin{enumerate}
\item Define a type corresponding to the statements of the form  "the $n$-ary function $f$ is primitive recursive ''.
\item Prove handy lemmas which may help to prove that a given function is primitive recursive. These lemmas can be considered as a way to build \emph{silently} large terms of 
type (\texttt{PrimRec $n$}) in intermediate steps of the proof.
More, we can associate \emph{tactics} with these lemmas.

\end{enumerate}

@@

The library \texttt{primRec} contains a large catalogue of lemmas allowing to prove statements 
of the form (\texttt{isPR $n$ $f$}). We won't list all these lemmas here, but give a few examples of
how they may be searched, then applied.

\begin{remark}
In the library \texttt{primRec}, all these lemmas are opaque (registered with \texttt{Qed}). Thus they do not allow the user to look at the witness of a proof of a \texttt{isPR} statement. 
It may be useful to make transparent all the instances of
\texttt{isPR} in the \texttt{Ackermann} and \texttt{goedel} libraries.
\end{remark}

@@

Simple proofs of statements (\texttt{isPR $n$ $f$}) may be just 
applications of the constructor \texttt{is\_pr}, often
thanks to the tactic call \texttt{exists $x$} where $x$ 
is some (hopefully) correct term of type (\texttt{PrimRec $n$}).

Let us show a few examples from
\href{../theories/html/hydras.MoreAck.PrimRecExamples.html}%
{Ackermann.MoreAck.PrimRecExamples.v}\footnote{Some of them are also in \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.}.

@@

\input{movies/snippets/primRecExamples/SuccIsPR}

@@

Please note that the projection $\pi_{1,1}$ is just the identity on \texttt{nat}, and is realized by 
(\texttt{projFunc 1 0}) (see Sect.~\vref{sect:primrec-examples2}).

@@

\subsubsection{The predecessor (total) function}

The predecessor function is defined by the following equations:

\begin{align*}
\texttt{pred}(0) &= 0 \\
\texttt{pred}(1+n) &= n = \pi_{1,2}(n,\texttt{pred}(n))
\end{align*}

This definition is easy to translate into a term of type \texttt{PrimRec 1}.

\inputsnippets{PrimRecExamples/xpred}

@@

Let us look at the proof that any constant $n$ of type \texttt{nat} has type (\texttt{PR 0})
(lemma  \texttt{const0\_NIsPR} of \texttt{primRec}). We carry out a proof by induction on $n$, the base case of which is already proven.
Now, let us assume $n$ is \texttt{PR $0$}, with $x:\texttt{PrimRec}\,0$ as a ``realizer''.
Thus we would like to compose this constant function with the unary successor function.

@@

\vspace{4pt}
Here is a quite simple proof of \texttt{const0\_NIsPR}.

@@

\subsubsection{Another proof that \texttt{Nat.add} is primitive recursive}

We have already proven that \texttt{Nat.add} is primitive recursive. The following alternative proof, --- more detalled  ---, 
shows how to search and apply lemmas from
the \texttt{Ackermann} library.

Let us look for some lemma which could help to prove some 
recursive arithmetic binary function is primitive recursive.

\inputsnippets{PrimRecExamples/PRnatRecSearch}

Good! Let us express addition in terms of \texttt{nat\_rec}.

\inputsnippets{PrimRecExamples/plusIsPRa}

The lemma \texttt{Ackermann.isPRextEqual} tells us that if a function $g$ is extensionally equal to a primitive recursive function, then $g$ is primitive recursive too.

\inputsnippets{PrimRecExamples/isPRextEqual}

% The lemma \texttt{plusIsPR} is already proven in \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}. We present in 
% \href{../theories/html/hydras.MoreAck.PrimRecExamples.html}{MoreAck.PrimRecExamples}
% a commented version of this proof,

% First, we look for lemmas which may help to prove that a given function obtained with the recursor \texttt{nat\_rec} is primitive recursive.

% \input{movies/snippets/PrimRecExamples/PRnatRecSearch}

% Thus, let us define an helper and prove its equivalence with \texttt{Nat.add}.

% \input{movies/snippets/PrimRecExamples/plusIsPRa}

@@

Let us start our proof.

\inputsnippets{PrimRecExamples/plusIsPRa1,
  PrimRecExamples/plusIsPRa2}

@@

\vspace{4pt}
\input{movies/snippets/PrimRecExamples/plusIsPRa3}
\vspace{4pt}

Thus, our first subgoal is easily solved. The rest of the proof 
is just an application of already proven lemmas.

@@

\index{ackermann}{Exercises}
\begin{exercise}
A few lemmas similar to \texttt{filter010IsPR}, also shown  in the \texttt{primRec} library help the user to control the arity of functions.
Thus, the reader may look at them, and invent h.er.is simple examples of application.
\end{exercise}

@@

@@

\subsection{More advanced examples}

\subsubsection{The minimum of two natural numbers}

Let $a$ and $b$ be two natural numbers. The \emph{minimum} of $a$ and $b$ is $a$ if $a \leq b$, otherwise $b$.

Thus, we propose the following definition:

\inputsnippets{MinPR/minAltDef}

Here is a sketch of proof that standard library's \texttt{min} is
primitive recursive. The reader is kindly invited to fill the missing steps.

\inputsnippets{MinPR/minProof1, MinPR/minProof2, MinPR/minProof3}
